Zohmg---a large scale data store for aggregated time-series-based data. [DRAFT]

### 2009-05-18 doc/report/msc-report.tex
doc/report/msc-report.tex: * checked a lot of dave's comments. * added apache log example, maybe needs appropriate placing. * work on implementation chapter (WIP).

@@

\section{New techniques}

@@

BigTable is a data store designed for traversal of very large data sets.
Time-stamped cells inside column-families in rows. The row has a key and any
number of cells attached to it. The rows in a table are sorted by the key.
The typical use-case of BigTable is to scan a range of rows. It is therefor
important to set up the row keys so that related data is close to each other.
For example, if it makes sense to traverse the data in chronological order, the
key might contain a representation of the data's timestamp. The classic example
is to have the reversed domain name (i.e. com.google.www) as the row key, which
means that all sub-domains of any domain are next to each other.

\subsubsection{Data model}

The data model is a simple one: Rows of data are stored in labeled tables. Each
row is indexed with a row key, a column key, and a timestamp. The columns are
of the form "column-family:qualifier", where the column-family is one of a
number of fixed column-families defined by the table, and the qualifier is an
arbitrary string specified by the application. In effect, column families are
pre-announced when creating the table while qualifiers are not. The contents of
each column family are stored together, so the user will want to store items
that have similar characteristics in the same column family.  Each data point is
called a cell. A cell is an uninterpreted array of bytes, there are no data
types, and it is up to the application to interpret the data correctly. The
table can store any number of timestamped versions of each cell, allowing
versioning of data.

The model above can indeed be thought of as a sparse distributed
multidimensional sorted map. In concept, this multidimensional map is identical
to a nested hash table. The dimensions of this multidimensional map is mapped
out on the row key, the column family and the column qualifier, and possibly
also of the version timestamp.

@@

@@

\section{System function}

@@

TODO: How to discuss modelling of data in a general sense, without mentioning
MapReduce and BigTable?
% thinking in simple data structures,
% ..

@@

\section{System requirements}

@@

\subsection*{Exploration by prototyping}

@@

\subsubsection*{Literature studies}
% vet inte riktigt vad jag vill saga om det har.

@@

\subsubsection*{Python}

@@

Because there existed a Python module for writing MapReduce programs (Dumbo,
further later in this chapter), Python was a natural choice during the design
and exploration phases. The main reason was the rapid development of Python
programs, several times shorter than their Java counter parts made them faster
to develop.

@@

@@

Every Hadoop node can be formatted with HDFS, this reserves disk space on
the node to be used for HDFS. The HDFS container resides on the nodes
general purpose file system. It is also possible to use HDFS as a
stand-alone general purpose distributed file system, without Hadoop.

@@

% TODO> why is this Tool & BigTable-subsection called Dimensions?
% Disambiguation needed. -fredrik
\subsection*{Physical storage}

Each column-family is stored as a single file in HBase; it is then subdivided
into column-qualifiers within the same file. This is important to consider when
designing application and schemas, since reading from several column-families
will open several files.

@@

Depending on how the row keys are formatted different, hierarchies are created.
The order in which the data is stored is important, it defines how effective
scanning data will be. Sparsity also comes into play concerning effective
scanning. In the case of having thousand rows and requesting ten out of these,
then only one por cent of the data is interesting. If a scan would be required
to visit all the thousand rows a lot of rows are skipped. Skipping rows is
expensive in that sense that they are but the result is thrown away. The goal is
to push this cost to a bare minimum, using as many of the rows as possible.

@@

Thrift is a remote procedure call framework for building scalable cross-language
services. Thrift combines a software stack with a code generation engine to
build services that work seamlessly between a wide range of programming and
scripting languages.

Thrift was originally developed at Facebook and in April 2007 it was released as
free software. It entered the Apache Incubator in May 2008. \cite{thrift}

@@

While relational databases can be made highly available and reliable with load
balancing and so called \textit{Sharding}, the setup is non-trivial. Regarding
scalability, relational databases are highly dependent on having data in working
memory; thus they scale poorly.
% TODO: [citation needed]

@@

\section{Project management}

@@

The environment is a Python script which defines run-time variables, required by
Hadoop and Dumbo. Using a Python scripts for run-time variables leaves the
problem of validation to Python. The environment configuration file resides in
the project's \texttt{config} directory.

@@

Going back to the running example of apache access logs, a usermap function
might look something like this:

@@

The actual parsing of the log is abstracted; in the above example the
emitting of dimensionality is showcased. It is assumed that
\texttt{parse\_apache\_log()} accepts a single line from the log file and
returns an object with the appropriate fields, such as useragent and
statuscode.

For every line of the log file, a point in n-space (the dimensions and the
timestamp) is emitted together with a count of pageviews and bytes; one
pageview per request and the number of bytes served in the request. It is
possible to add business logic here, perhaps only counting requests from
useragents that are not web spiders and that resulted in a HTTP 200 status
code.

The last line \textit{yields} the tuple. This is a Python construct called a
\textit{generator} that lets the map function emit any number of tuples (or
none!) for a single invocation.

The user will import a day worth of logs like so:

@@

% TODO: method.
-- Limits on cardinality --

@@

@@

\section{Data interchange}

\subsection{Internal}

Typed bytes

\subsubsection{Input formats}

@@

\subsubsection{Output formats}

% TODO: not truly correct
The Darling output format for HBase. Between Zohmg and Hadoop JSON is used
for serializing data.

\section{Exporting data}

@@

When a user queries the data server, JSON\footnote{JavaScript Object Notation}
is returned. JSON was chosen because it is i simple text serialization format
which is widely used in web development.

\subsection{External}

@@

The URLParser class is middleware that comes with Python Paste; its purpose is
to dispatch to applications based on the URL. The URLParser removes the first
part of the URL and tries to find a matching application to relay the request
to. For instance, consider the URL \texttt{/client/file.html}. The first
part---\texttt{client}---will be removed and the request \texttt{/file.html}
will be sent to the application named \texttt{client.py}.

@@

Fetches data from HBase and transforms it with the requested transform
program from the project's transformers directory, output is
dumped as JSONP.

@@

\section{Future work}

@@

\subsection{Multiple data sets}

@@

\subsection{Metadata}

@@

\subsection{Updating data set files}

@@

\subsection{Importing without MapReduce}

@@

Install Hadoop and HBase.

@@

>>>>

% TODO: placeholder quote.
\chapterquote{One day some of the kids from my neighbourhood carried my mother's
groceries all the way home---you know why? It was out of respect!}{Travis
Bickle}

@@

\section{New Techniques}

@@

BigTable is a data store designed for traversal of very large data sets.
BigTable's cells are basically indexed by row key, column key, and a time-stamp.
Each row has a key and any number of cells attached to it. The rows in a table
are sorted by their keys. The typical use-case of BigTable is to scan a range of
rows. It is therefor important to set up the row keys so that related data is
close to each other. For example, if it makes sense to traverse the data in
chronological order, the key might contain a representation of the data's
timestamp. The classic example is to have the reversed domain name (i.e.
com.google.www) as the row key, which means that all sub-domains of any domain
are next to each other (see table 2.1).

\subsubsection{Data Model}

BigTable's data model can indeed be thought of as a sparse distributed
multidimensional sorted map. In concept, this multidimensional map is identical
to a nested hash table. The dimensions of this multidimensional map is mapped
out on the row key, the column family and the column qualifier, and possibly
also of the version timestamp.

@@

The data model is a simple one: Rows of data are stored in labeled tables. Each
row is indexed with a row key, a column key, and a timestamp. The columns have
the form "column-family:qualifier", where the column-family is one of a number
of fixed column-families defined by the table's schema, and the qualifier is an
arbitrary string specified by the application. In effect, column-families are
pre-announced when creating the table while qualifiers are not. The contents of
each column family are stored together, so the user will want to store items
that have similar characteristics in the same column family.  Each data point is
called a cell. A cell is an uninterpreted array of bytes, there are no data
types, and it is up to the application to interpret the data correctly. The
table can store any number of timestamped versions of each cell, allowing
versioning of data.

@@

\section{System Function}

@@

TODO: How to discuss modelling of data in a general sense, without mentioning
MapReduce and BigTable?
% thinking in simple data structures,

@@

\section{System Requirements}

@@

\subsection*{Exploration}

The possible tools were explored by testing and benchmarking them.

The problem area was explored by prototyping solutions to a given subproblem.

\subsubsection*{Software}

% TODO: renskriv.
MapReduce frameworks: Hadoop, disco.

Hadoop is used in production at Last.fm, it made sense to use it because the
work done would integrate with the Last.fm's infrastructure easily.

There exists a few databases which inspired by BigTable; they aim to be
distributed and have column-oriented properties: HBase, Cassandra, MonetDB,
MongoDB, Hypertable. CouchDB is an document-oriented database.

These databases were explored and tried out. Sadly most of these BigTable-like
databases were in a very early phase of development and hard to even install.
The experience was that HBase was the most mature of them. Maturity was the
compound experience with the software: community, codebase, documentation,
stability etc. HBase is also tightly coupled with Hadoop which is used in
production at Last.fm.

CouchDB was experienced to be mature as well but lacked integration with
MapReduce, all database communication would be necessary to do over HTTP.
Setting up CouchDB with replication required some hands-on, it lacks sorting of
documents which BigTable-like databases utilize.

\subsubsection*{Benchmarks}

Only HBase made it this far, the other databases in the roster was hard or
impossible to either compile, install, setup, or all of them.

HBase was benchmarked (the benchmarks was similar to the ones found in the
BigTable paper?).

\subsubsection*{Prototyping}

@@

\subsection*{Literature studies}
% vet inte riktigt vad jag vill saga om det har.

@@

\subsection*{Python}

@@

Because there existed a Python module for writing MapReduce programs (Dumbo for
Hadoop, both further later in this chapter), Python was a natural choice during
the design and exploration phases. Python was also a good choice because of the
rapid prototype development it enables.

@@

% TODO: method.
TODO: -- Limits on cardinality --

@@

Every Hadoop node can be formatted with HDFS, this reserves disk space on the
node to be used for HDFS. The HDFS partition\footnote{The HDFS partition is a
directory containing HDFS files on a general-purpose file system.} resides on
the node's general purpose file system. It is also possible to use HDFS as a
stand-alone general purpose distributed file system, without Hadoop.

@@

\subsection*{Physical storage}

Each column-family is stored as a single file in HBase; it is then subdivided
into column-qualifiers within this file. This is important to consider when
designing application and schemas, since reading from several column-families
will open several files.

@@

Depending on how the row keys are formatted different, hierarchies are created.
The order in which the data is stored is important, it defines how effective
scanning data will be. Sparsity also comes into play during effective scanning.
In the case of having thousand rows and requesting ten out of these, then only
one per cent of the data is interesting. If a scan would be required to visit
all the thousand rows a lot of rows are skipped. Skipping rows is expensive in
that sense that they are visited but the result is thrown away. The goal is to
push this cost to a bare minimum, using as many of the rows as possible.

@@

Thrift is a remote procedure call framework for building scalable cross-language
services. Thrift combines a software stack with a code generation engine to
build services that work seamlessly between a wide range of programming and
scripting languages. \cite{thrift}

Thrift was originally developed at Facebook and in April 2007 it was released as
free software. It entered the Apache Incubator in May 2008.

@@

While relational databases can be made highly available and reliable with load
balancing and so called \textit{sharding}, the setup is non-trivial. Regarding
scalability, relational databases are highly dependent on having data in working
memory; thus they scale poorly.
% TODO: [citation needed]

@@

\section{Project Management}

@@

The environment is a Python script which defines run-time variables, required by
Hadoop and Dumbo. Using a Python script for run-time variables leaves the
problem of validation to Python. The environment configuration file resides in
the project's \texttt{config} directory.

@@

% TODO: move apache logs to introduction perhaps.
An example data source is the Apache HTTP Server access logs. The Apache HTTP
Server Combined Log Format is commonly used to log requests. The log format is
configurable, an example log line might have the following format

\begin{lstlisting}[caption=Example log line entry for the Apache HTTP Server's
Combined Log Format.,captionpos=b]
   127.0.0.1 - emma [20/Apr/2009:13:37:00 +0100]
   "GET /music HTTP/1.0" 200 2326 "http://last.fm/"
   "Mozilla/5.0 (X11; U; Linux i686)"
\end{lstlisting}

The above example log entry is line-broken due to space constraints, normally
the entry would be on one line. The example log line contains access info for a
specific request; it contains the key entries: the client's IP address,
timestamp, requested resource, HTTP status code for the request, the client's
user agent etc.

Revisiting the running example of Apache access logs, a usermap function might
look something like this:

@@

The above example showcases the emitting of dimensionality. In the example the
actual parsing of the log is abstracted out by the function
\texttt{parse\_apache\_log()}; it is assumed that this function accepts a single
line from an Apache HTTP Server log file and returns an object with the
appropriate fields, such as useragent and statuscode.

For every line of the log file, a point in n-space (the dimensions and the
timestamp) is emitted together with a count of pageviews and bytes; one
pageview per request and the number of bytes served in the request. It is
possible to add business logic here, perhaps only counting requests from
useragents that are not web spiders which resulted in a HTTP 200 status code.

The last line \textit{yields} the tuple. This is a Python construct, similar to
an iterator, called a \textit{generator} that lets the map function emit any
number of tuples for a single invocation.

The user will import a day worth of logs by invoking

@@

@@

\section{Data Interchange}

@@

\subsection*{Internal}

Using JSON for data interchange between map and reduce parts in Zohmg import
jobs.

For better performance typed bytes may be used. This is what Dumbo uses
internally.

\subsubsection{Input Formats}

An input format is a Java class which Hadoop uses for interpreting the input
data.

TableInputFormat

@@

\subsubsection{Output Formats}

% TODO: not truly correct
The Darling output format for HBase. Between Dumbo, which Zohmg uses for data
import, and Hadoop JSON is used for serializing data.

The Darling output format increased import speed by a factor ten, compared to
using Thrift.

\section{Exporting Data}

@@

When a user queries the data server, JSON\footnote{JavaScript Object Notation}
is returned. JSON was chosen because it is i simple text serialization format
which is widely used in web development. The Python module simplejson was
selected for the this implementation detail. \cite{simplejson}

\subsection*{External}

@@

The URLParser class is a middleware application that comes with Python Paste;
its purpose is to dispatch to applications based on the URL. The URLParser
removes the first part of the URL and tries to find a matching application to
relay the request to. For instance, consider the URL \texttt{/client/file.html}.
The first part---\texttt{client}---will be removed and the request
\texttt{/file.html} will be sent to the application named \texttt{client.py}.

@@

The transformers application fetches data from HBase with the same mechanism as
the raw data application. The data is transformed with a transformer---a Python
program in the project's \texttt{transformers} directory---requested by the
user. Similar to the raw data serving, the output is served as JSONP.

@@

\section{Future Work}

@@

\subsection*{Multiple Data Sets}

@@

\subsection*{Metadata}

@@

\subsection*{Updating Data Set Files}

@@

\subsection*{Importing Without MapReduce}

@@

Hadoop and HBase is required to run Zohmg, they form the foundation of data
processing and storage which Zohmg builds upon. Zohmg is built with Python and
also requires a few Python modules, on a Debian based system these are installed
by the install script (see below).

To install Hadoop and HBase automatically a script is provided.

@@

The Zohmg installation depends on Python Setuptools \cite{setuptools}, which
needs to be installed prior to the installation.
